Verifying synchronous distributed algorithms using IVy

### 2018-08-22 paper/paper.tex
implented comments2

  Ivy is tailored to simplify deductive verification of parameterized distributed algorithms. There are a variety of systems currently implementing the deductive verification process. IVy makes use of the same underlaying principle, but it aims to increase user friendliness of the verification process.  IVy's hypothesis is that ``automated methods are difficult to apply in practice not primarily because they are unreliable, but because they are opaque''; meaning that they fail in ways inexplicable to the user. In light of this, the IVy system was designed to display proof-failure comprehensively to the human user, making the root cause of errors more visible. \cite[ p.1]{ivy}

@@

\section{IVy}
IVy is a system designed to verify correctness of infinite-sate systems using the aforementioned deductive approach. In a nutshell, it takes as input the specification of an algorithm, which may mutate state, and a set of invariants. It proves the inductiveness of the invariants in respect to the algorithm, or displays a counterexample to induction. If the set of invariants entails the desired properties of the algorithm, and Ivy proved induction, then it was proven that the algorithm is \textit{correct}.

In Ivy, the entity ``calling'' the algorithm in different state configurations is called the ``environment''. The environment conceptually represents the final users of the algorithm. An algorithm or protocol is \textit{correct} or \textit{safe} if it cannot be called by the environment in a state configuration that causes some pre-defined properties (assertions, invariants) to be violated. IVy supports verification of systems that have multiple methods (\textit{actions}) exported to the environment. For example to verify servers that specify connection and disconnection algorithms. This can be thought of as proving inductiveness of two independent algorithms with respect to the same invariants and system-state.

@@

\subsubsection{Step 1 - encoding in IVy language}
Step 1 requires the user to understand the Ivy language. This language is a specification language tailored to describe state changes in a system. Ivy language code is inherently different from procedural languages, as it serves a different purpose and its logic is never executed. The Ivy documentation itself calls the language features "unusual" \cite{refLanguageDoc} so a user may need considerable effort to understand how to best use its tools.

Using their understanding of the Ivy language users have to encode their algorithm in it. However, distributed algorithms rarely exist in a vacuum. Algorithms such as \textit{Floodset} are defined within a \textit{system context} or \textit{framework} \cite{refNancy}, and make no sense without first defining this \textit{framework}. Thus users have to encode their algorithm together with the \textit{framework} in which the algorithm was designed to execute. This may include things such as network topology, the amount of processors that may fail, how and when processors fail, details about message sending, etc.

@@

In Ivy one often wants to specify system-wide state-changes. For this reason Ivy allows iterator-free assignments involving all instances of a type. To make such an assignment one can use placeholders that are denoted by uppercase tokens. For example, to set all nodes without an incoming message to \textit{failed}:

@@

\textit{Assume} is used to exclude the possibility of something happening.  This is mostly useful in combination with non-deterministic assignments that one wants to restrict. The following action chooses a random node to assign to \textit{y}, but will not allow control to pass if the chosen node has failed. Therefore the actions will choose a random non-failed node to send a message to x.  If there is no node that has not failed, this action cannot terminate.

\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
action random_node_sends_message_to(x:node) = {
  y := *;
  assume ~failed(y);
  send(x, y);
}
\end{lstlisting}
\end{mdframed}

\textit{Assume} simplifies modelling in some situations, as it can easily eliminate impossible behaviour. However,
it is risky to use \textit{assume} for the same reason: one might dismiss a valid counterexample to the specification
by assuming it cannot happen. To give an extreme example, putting $assume\ false$ at the beginning of an exported action will cause all of Ivy's checks for induction to come out positive.

@@

\subsection{Decidability}
First order logic is undecidable in general, but still many interesting properties can be expressed in a subset of first oder logic called the ``decidable fragment''. Ivy's documentation defines this subset of first-order formulas. \cite{decid}

For example, Ivy only allows stratified functions. Meaning that a graph, with every type being a node and every function mapping from type $A$ to type $B$ be an edge from $A$ to $B$, has to be cycle free. Thus, there cannot be a function from $X \rightarrow Y$ and from $Y \rightarrow X$ in the same specification.

@@

While a complete understanding of decidability in Ivy is advantageous, to use Ivy successfully it is not strictly necessary. This is because Ivy assists the user and tries to automatically determine if formulas lie outside of the decidable fragment. It warns users about specifications that it estimates are undecidable.  It must be noted that IVy's check is conservative as ``fully unfolding all of the definitions could be exponential''. Ivy may report specifications as undecidable which are in fact decidable. For this reason IVy allows the check to be deactivated. \cite{decid}

@@

The algorithm that was implemented in IVy as part of this work is called \textit{Floodset} and was taken out of Lynch's book ``Distributed Algorithms'' \cite{refNancy}.
\textit{Floodset} is a consensus algorithm for synchronous networks with failing processes.
To understand \textit{Floodset}, one needs to first understand the aforementioned \textit{framework}/\textit{system} within which this algorithm operates, which is also defined in \cite{refNancy}.

@@

The network of processes is represented as a directed graph $G=(V,E)$. Where $n$ is $|V|$, the number of nodes in the network. Each node represents a process in the network.
The communication paths between nodes are represented by directed edges in the graph.
$outnbrs_i$ is the set of nodes to which an edge from node $i$ exists.
$innbrs_i$ is the set of nodes from which an edge to node $i$ exists.
$M$ denotes the message alphabet, meaning that elements of $M$ can be used to construct messages (vectors of elements of $M$) from nodes to other nodes.
There are a number of components associated with each node:
\begin{itemize}
    \item $states_i$, a set of states (can be infinite)
    \item $start_i$, a nonempty subset of $states_i$
    \item $msgs_i$, a message generation function that maps $states_i \times outnbrs_i$ to elements of $M \cup \{null\}$
    \item $trans_i$, a state transition function that maps $states_i \times vectors\ of\ M$ index by $innbrs_i$ to $states_i$
\end{itemize}
Thus, a node is always in some state in $states_i$. Importantly, the set of states is not necessarily finite, meaning that unbounded data structures like arrays can be modeled in the system.
Each node defines a message generation function $msgs_i$, that uses the current state to return a message for each out neighbour. Similarly, each node has a state transition function $trans_i$, that is used to determine
from the current state and all incoming messages the next state the node will transition to.

@@

\noindent \textbf{Lemma 1}:\\
If no process fails during a particular round $r,\ 1 \le r \le f + 1$, then $W_i(r) = W_j(r)$ for all $i$ and $j$ that did not fail after $r$ rounds.\\
\textbf{Proof}:\\
\textit{Indirect}: If there exists a value $v \in V$ that is in $W_i(r)$ but not in $W_j(r)$ after round $r$, in which no process failed, that implies, that some node $k$ sent a value to $i$ that it did not send to $j$, or $i$ failed to send a member of $W_i(r-1)$. Since all nodes broadcast their $W$, this is only possible if node $k$ or $i$ failed to send at least one message. But since no processes failed, each process must have shared their $W$ with every other process. \textit{Contradiction}. \\

@@

\noindent \textbf{Lemma 3}:\\
After $f+1$ rounds there has been a round $r'$ in which no process failed\\
\textbf{Proof}:\\
\textit{Indirect:} If a process failed in all of $f+1$ rounds at least on process must have failed in all of the $f+1$ rounds. This implies
that at least $f+1$ processes failed. Only $f$ processes can fail. \textit{Contradiction}.\\

@@

\noindent The Lemmas allow proving the theorem:\\
\noindent \textbf{Theorem}:\\
\textit{Floodset} satisfies the correctness conditions.\\
\textbf{Agreement}:\\
From \textit{Lemma 4} we know what $W_i = W_j$ after $f+1$ rounds. Since all decision conditions are the same and only depend on $W$, agreement is satisfied.\\
\textbf{Validity}:\\
If all processes start with the same initial value $v \in V$, then there exists no value $v', v \neq v'$, that is part of any $W$. Thus, from $W_i(f+1) = W_j(f+1)$, $\neg failed(i) \land \neg failed(j)$, and $W_i(f+1) = \{v\}$ it follows
that every active node decides $v$.\\
\textbf{Termination}:\\
Trivially, nodes decide after $f+1$ rounds.\\

@@

\section{Specification}
To summarize: we have a parameterized system with $n$ process and $f$ number of faults. We want to show \textit{Floodset} is correct regardless of the parameters.
It is trivial to verify for a small number of $n,\ f$, but the question is ``does \textit{Floodset} work for any assignment of $n,\ f,\ n < f$''.
To answer this question the algorithm was encoded in the IVy language and then inductive invariants were constructed to prove correctness.

@@

The assumption $f<n$, ``number of failed nodes is bounded to an amount smaller than the total number of nodes'', can alternatively be expressed via $\exists n\ \neg failed(n)$. This is because the existence of a functioning node implies that the number of failed nodes $f$ is smaller than $n$.

@@

\subsection{Types, functions, relations}
First the types, functions, relations and individual variables that are used throughout the program are presented. They are annotated with comments explaining their purpose.  Their meaning will become more clear as the code is presented in the following paragraphs.

@@

\noindent \textit{round} is the only exported action. It specifies the basic structure of Lynch's \textit{framework}.
\noindent Line 2 makes it illegal for the environment to call \textit{round} if decisions already occurred.
\noindent Line 3 does some bookkeeping.
\textit{call step1} models the communication step. \textit{simulate\_failure} simulates failure before, during
and after \textit{step1}. After \textit{decision\_round} is set, \textit{call step2} is executed to model the state transition of the nodes.

@@

The actual delivery of a message is modeled via the \\
$incoming\_message(RECIPIENT:node,SENDER:node)$ relation. Where $incoming\_message(0,1) = true;\ incoming\_message(0,2) = true;$ means that node $0$ receives the messages from node $1$ and node $2$.

These relations allow us to use IVy's assignments to easily model \textit{Step 1} in a round of \textit{Floodset}.
Line 2 assigns for all nodes $A$ and values $V$ to $message(A,V)$ the boolean value of $node.w(A,V) \land \neg failed(A)$. In other words, every node that has not failed prepares a message with all values in its $w$. Line 3 states that all nodes that haven't failed send messages to all other nodes.

@@

This section shows how the stopping failure model described by Lynch \cite{refNancy} and explained in \textit{section III ii.} was modeled in Ivy.

@@

Lines 2 to 5 resemble the snippet of section \textit{vii.1}. In line 2 we choose a random number of nodes to fail. The \textit{assume} in line 3 then excludes all cases where a node that already failed was chosen to fail again. Line 4 adds the newly failed nodes to the set of failed nodes. Line 5 enforces $f < n$. Since this action is only called once per round we do not need an extra relation \textit{failing\_now} presented in \textit{vii.1}.

@@

\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
invariant system.decision_round <->
            (forall N . (~node.failed(N) ->
                          (exists V . node.decision(N,V))))
\end{lstlisting}
\end{mdframed}
This invariant assumes that the system will eventually reach the decision round. However, truly verifying \textit{termination} would require checking that the system must reach the decision round, which means verifying  \textit{liveness}. Ivy theoretically allows \textit{liveness} checks by constructing a reduction of \textit{liveness} to \textit{safety} \cite{live} but this is out of the scope of this work.

\subsection{Invariants for Inductiveness}
If one were to run $ivy\_check$ on the program without specifying any
additional invariants, IVy finds that the \textit{agreement} invariant is not inductive,
and produces a counter example to induction (CTI). The CTI can be drawn in a GUI.

@@

\noindent In the left hand section of the program we can see, that the system transitioned from state 0
to state 1 by executing \textit{round}. In the center IVy displays state 0, meaning the state
before \textit{round} was executed. The brown edges specify the relation $w$, the purple edges
show $val$.

State 0 shows, that no node has their initial value, which is modeled by the relation $val$,
in its $w$. The user must now realise, that this is an unreachable state, and exclude it by coming
up with an invariant.

@@

The method for finding the remaining invariants is identical: Execute $ivy\_check$ $\rightarrow$ examine CTI $\rightarrow$ design invariant $\rightarrow$ execute $ivy\_check$. The following snippets presents the necessary invariants for induction annotated with comments.
\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
# the global default is the one the nodes know
invariant node.default(N) = node.def

@@

@@

Since any check on \textit{Floodset}'s correctness assumes that the \textit{framework} has been modeled correctly, it makes sense to include correctness conditions specifically aimed at the \textit{framework} in the form of invariants. These invariants are not necessary for induction. Removing them will not cause any of Ivy's other checks to fail. Their sole purpose is to give the user confidence that the \textit{framework} is specified correctly.

@@

\section{Evaluation}
How viable is IVy as a tool for algorithm designers proving correctness
of their work? The goal, when turning to a tool like IVy, is certainty of
correctness. However, due to a variety of reasons, absolute certainty is never achieved.
This section evaluates Ivy's usefulness pointing out strengths and limitations that Ivy users should keep in mind
when working with the system.

\subsection{Equivalence of specification and algorithm}
One can not trivially determine, that the specification in the IVy language is logically equivalent to the actual algorithm the user wishes to verify. Whether the algorithm to be verified exists in the user's mind, in the form of pseudo-code or fully implemented in a programming language, no guarantee can be given in regards to the correctness of the translation of the algorithm into the Ivy language. Some examples illustrating this difficulty will be given in the following paragraphs.

\subsubsection{Structural difference of Ivy and procedural languages}
The IVy language is structurally unlike classical procedural languages. This makes it inherently challenging to achieve a one-to-one translation from a classical distributed algorithm, that was specified in a procedural way, like \textit{Floodset}. An example of this structural difference is the relatively complicated quantifier-logic in assignments that update the global network state. Take as an example the first line of \textit{step2}:

@@

This comparison clearly shows the complex structural differences. While the pseudo-code is concerned only with state alteration of a single node, the Ivy language code simultaneously alters the state of all nodes. It uses an existential quantifier nested within a term. Its logic touches on multiple layers of abstraction as it mixes system and node concerns. It considers node failure, message reception and node state alteration.

As these two lines are so different, equivalence may not be obvious to a user. Proving equivalence of pseudo-code and Ivy language specification would be a non-trivial proof exercise that certainly trumps manually proving \textit{Floodset}'s correctness in difficulty.

@@

\subsection{Accidental invalidation of legal state configurations with \textit{assume}}
$assume$ statements may silently invalidate state configurations. They therefore pose a threat as they
may allow false-positive Ivy checks.
To give a practical example, the author of this paper originally modeled the line

@@

\noindent For example, one may insert
\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
assert ~(exists N . node.failed(N) &
        (exists J . ~incoming_message(J,N) &
          exists K . incoming_message(K,N) & J~=N & K~=J & N~=K))
\end{lstlisting}
\end{mdframed}
after the \textit{simulate\_failure} action. The negated part, $\gamma$, specifies that there exists a node $N$, that failed and failed to deliver a message to $J$ while successfully delivering a message to $K$, with $N,J,K$ distinct. By negating $\gamma$, we force Ivy to construct a counterexample such that $\gamma$ is satisfied. If IVy does not find a counterexample to this assertion, it means that $\gamma$ cannot occur. In this particular case, that would mean the specification does not allow partial message delivery. The specification would therefore be erroneous. One has to finally remove the failing assertion to allow Ivy to prove induction again.

@@

\subsection{Necessity of deep understanding}
It has become apparent, that successful verification with IVy is not possible without deep knowledge about how the algorithm functions. Ivy is not a ``translate and go'' solution.
Evidence of this include the complex invariant that requires knowledge about \textit{lemma 1} or the alternative encoding for $f<n$ and the decision condition $if\ rounds = f+1$.

>>>>

  IVy is tailored to simplify deductive verification of parameterized distributed algorithms. There are a variety of systems currently implementing the deductive verification process. IVy makes use of the same underlaying principle, but it aims to increase user friendliness of the verification process.  IVy's hypothesis is that ``automated methods are difficult to apply in practice not primarily because they are unreliable, but because they are opaque''; meaning that they fail in ways inexplicable to the user. In light of this, the IVy system was designed to display proof-failure comprehensively to the human user, making the root cause of errors more visible. \cite[ p.1]{ivy}

@@

\section{IVy}
IVy is a system designed to verify correctness of infinite-sate systems using the aforementioned deductive approach. In a nutshell, it takes as input the specification of an algorithm, which may mutate state, and a set of invariants. It proves the inductiveness of the invariants in respect to the algorithm, or displays a counterexample to induction. If the set of invariants entails the desired properties of the algorithm, and IVy proved induction, then it was proven that the algorithm is \textit{correct}.

In IVy, the entity ``calling'' the algorithm in different state configurations is called the ``environment''. The environment conceptually represents the final users of the algorithm. An algorithm or protocol is \textit{correct} or \textit{safe} if it cannot be called by the environment in a state configuration that causes some pre-defined properties (assertions, invariants) to be violated. IVy supports verification of systems that have multiple methods (\textit{actions}) exported to the environment. For example to verify servers that specify connection and disconnection algorithms. This can be thought of as proving inductiveness of two independent algorithms with respect to the same invariants and system-state.

@@

\subsubsection{Step 1 - encoding in IVy language}
Step 1 requires the user to understand the IVy language. This language is a specification language tailored to describe state changes in a system. IVy language code is inherently different from procedural languages, as it serves a different purpose and its logic is never executed. The IVy documentation itself calls the language features "unusual" \cite{refLanguageDoc} so a user may need considerable effort to understand how to best use its tools.

Using their understanding of the IVy language users have to encode their algorithm in it. However, distributed algorithms rarely exist in a vacuum. Algorithms such as \textit{Floodset} are defined within a \textit{system context} or \textit{framework} \cite{refNancy}, and make no sense without first defining this \textit{framework}. Thus users have to encode their algorithm together with the \textit{framework} in which the algorithm was designed to execute. This may include things such as network topology, the amount of processors that may fail, how and when processors fail, details about message sending, etc.

@@

In IVy one often wants to specify system-wide state-changes. For this reason IVy allows iterator-free assignments involving all instances of a type. To make such an assignment one can use placeholders that are denoted by uppercase tokens. For example, to set all nodes without an incoming message to \textit{failed}:

@@

\textit{Assume} is used to exclude the possibility of something happening.  This is mostly useful in combination with non-deterministic assignments that one wants to restrict. The following action non-deterministically chooses a node to assign to \textit{y}, but will not allow control to pass if the chosen node has failed. Therefore the actions will choose a non-failed node to send a message to x.  If there is no node that has not failed, this action cannot terminate.

\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
action some_node_sends_message_to(x:node) = {
  y := *;
  assume ~failed(y);
  send(x, y);
}
\end{lstlisting}
\end{mdframed}

\textit{Assume} simplifies modelling in some situations, as it can easily eliminate impossible behaviour. However,
it is risky to use \textit{assume} for the same reason: one might dismiss a valid counterexample to the specification
by assuming it cannot happen. To give an extreme example, putting $assume\ false$ at the beginning of an exported action will cause all of IVy's checks for induction to come out positive.

@@

\subsection{Decidability}
First order logic is undecidable in general, but still many interesting properties can be expressed in a subset of first oder logic called the ``decidable fragment''. IVy's documentation defines this subset of first-order formulas. \cite{decid}

For example, IVy only allows stratified functions. Meaning that a graph, with every type being a node and every function mapping from type $A$ to type $B$ be an edge from $A$ to $B$, has to be cycle free. Thus, there cannot be a function from $X \rightarrow Y$ and from $Y \rightarrow X$ in the same specification.

@@

While a complete understanding of decidability in IVy is advantageous, to use IVy successfully it is not strictly necessary. This is because IVy assists the user and tries to automatically determine if formulas lie outside of a decidable fragment. It warns users about specifications that it estimates are undecidable.  It must be noted that IVy's check is conservative as ``fully unfolding all of the definitions could be exponential''. IVy may report specifications as undecidable which are in fact decidable. For this reason IVy allows the check to be deactivated. \cite{decid}

@@

The algorithm that was verified in IVy as part of this work is called \textit{Floodset} and was taken out of Lynch's book ``Distributed Algorithms'' \cite{refNancy}.
\textit{Floodset} is a consensus algorithm for synchronous networks with failing processes.
To understand \textit{Floodset}, one needs to first understand the aforementioned \textit{framework}/\textit{system} within which this algorithm operates, which is also defined in \cite{refNancy}.

@@

The network of processes is represented as a directed graph $G=(V,E)$. The number $|V|$ of nodes in the network is denoted by $n$. Each node represents a process in the network.
The communication paths between nodes are represented by directed edges in the graph.
$outnbrs_i$ is the set of nodes to which an edge from node $i$ exists.
$innbrs_i$ is the set of nodes from which an edge to node $i$ exists.
$M$ denotes the message alphabet, meaning that elements of $M$ can be used to construct messages (vectors of elements of $M$) from nodes to other nodes.
There are a number of components associated with each node:
\begin{itemize}
    \item $states_i$, a set of states (can be infinite)
    \item $start_i$, a nonempty subset of $states_i$
    \item $msgs_i$, a message generation function that maps $states_i \times outnbrs_i$ to elements of $M \cup \{null\}$
    \item $trans_i$, a state transition function that maps $states_i \times vectors\ of\ M$ index by $innbrs_i$ to $states_i$
\end{itemize}
Thus, a node is always in some state in $states_i$. Importantly, the set of states is not necessarily finite, meaning that unbounded data structures like arrays can be modeled in the system.
Each node defines a message generation function $msgs_i$, that uses the current state to return a message for each out neighbour. Similarly, each node has a state transition function $trans_i$, that is used to determine
from the current state and all incoming messages the next state the node will transition to.

@@

\noindent \textbf{Lemma 1}:\\
If no process fails during a particular round $r,\ 1 \le r \le f + 1$, then $W_i(r) = W_j(r)$ for all $i$ and $j$ that did not fail after $r$ rounds.\\
\textbf{Proof}:\\
\textit{Indirect}: Suppose no process failed during round $r$, and there exists a value $v \in V$ that is in $W_i(r)$ but not in $W_j(r)$. This implies, that some node $k$ sent a value to $i$ that it did not send to $j$, or $i$ failed to send a member of $W_i(r-1)$. Since all nodes broadcast their $W$, this is only possible if node $k$ or $i$ failed to send at least one message. But since no processes failed, each process must have shared their $W$ with every other process. \textit{Contradiction}. \\

@@

\noindent \textbf{Lemma 3}:\\
After $f+1$ rounds there has been a round $r'$ in which no process failed.\\
\textbf{Proof}:\\
\textit{Indirect:} If a process failed in all of $f+1$ rounds at least on process must have failed in all of the $f+1$ rounds. This implies
that at least $f+1$ processes failed. Only $f$ processes can fail. \textit{Contradiction}.\\

@@

\noindent We now prove correctness of \textit{Floodset} by proving the three correctness conditions \textit{agreement}, \textit{validity}
 and \textit{termination}. \\
\noindent \textbf{Theorem}:\\
\textit{Floodset} satisfies the correctness conditions.\\
\textbf{Agreement}:\\
\textbf{Proof}:\\
From \textit{Lemma 4} we know what $W_i = W_j$ after $f+1$ rounds. Since all decision conditions are the same and only depend on $W$, agreement is satisfied.\\
\textbf{Validity}:\\
\textbf{Proof}:\\
If all processes start with the same initial value $v \in V$, then there exists no value $v', v \neq v'$, that is part of any $W$. Thus, from $W_i(f+1) = W_j(f+1)$, $\neg failed(i) \land \neg failed(j)$, and $W_i(f+1) = \{v\}$ it follows
that every active node decides $v$.\\
\textbf{Termination}:\\
\textbf{Proof}:\\
Trivially, nodes decide after $f+1$ rounds.\\

@@

\section{Specification}
To summarize: we have a parameterized system with $n$ process and $f$ number of faults. We want to show \textit{Floodset} is correct regardless of the parameters.
It is straightforward to check for a small number of $n,\ f$, but the question is ``does \textit{Floodset} work for any assignment of $n,\ f,\ n < f$''.
To answer this question the algorithm was encoded in the IVy language and then inductive invariants were constructed to prove correctness.

@@

The assumption $f<n$, ``number of failed nodes is bounded to an amount smaller than the total number of nodes'', can alternatively be expressed via $\exists n\ \neg failed(n)$. Intuitively, $f<n$ is true if and only if there is a process that has not failed. Or in other words, the existence of a functioning node implies that the number of failed nodes $f$ is smaller than $n$.

@@

\subsection{Types, functions, relations}
First the types, functions, relations and individual variables that are used throughout the program are presented.

\subsubsection{Types}
The types used are $node$ and $value$. The set of processes is represented by $node$, $value$ represents the set of values $V$. Both types are uninterpreted, meaning they represent an arbitrary set of instances with at least one element.  The cardinality of these sets is \textit{parameterized}, meaning IVy proves correctness for an arbitrary amount of $nodes$ and $values$.

\subsubsection{Functions}
The specification uses the function $val(N:node) : value$ to store the initial value of every node and $default(N:node) : value$ to represent each node's reference to the common default value.

\subsubsection{Relations}
The specification uses a variety of relations. E.g. to represent the set $W$, $w(N:node,V:value)$ is used. It specifies if a node $n$ has in its $W$ the value $v$. Another example is $message(N:node,V:value)$, which is used to specify the set of values any node is broadcasting in the current round.

The purpose of the remaining relations can be found in the comments above their definition in the code below. Their meaning will become clear as the code is presented in the following paragraphs.

@@

\noindent The only exported action is \textit{round}. It specifies the basic structure of Lynch's \textit{framework}.
\noindent Line 2 makes it illegal for the environment to call \textit{round} if decisions already occurred.
\noindent Line 3 does some bookkeeping.
\textit{call step1} models the communication step. \textit{simulate\_failure} simulates failure before, during
and after \textit{step1}. After \textit{decision\_round} is set, \textit{call step2} is executed to model the state transition of the nodes.

@@

The actual delivery of a message is modeled via the $incoming\_message(RECP:node,SEND:node)$ relation. Where $incoming\_message(0,1) = true;\ incoming\_message(0,2) = true;$ means that node $0$ receives the messages from node $1$ and node $2$.

These relations allow us to use IVy's assignments to easily model \textit{Step 1} in a round of \textit{Floodset}.
Line 2 assigns for all nodes $A$ and values $V$ to $message(A,V)$ the Boolean value of $node.w(A,V) \land \neg failed(A)$. In other words, every node that has not failed prepares a message with all values in its $w$. Line 3 states that all nodes that have not failed send messages to all other nodes.

@@

This section shows how the stopping failure model described by Lynch \cite{refNancy} and explained in \textit{section III ii.} was modeled in IVy.

@@

Lines 2 to 5 resemble the snippet of section \textit{vii.1}. In line 2 we choose a non-determined number of nodes to fail. The \textit{assume} in line 3 then excludes all cases where a node that already failed was chosen to fail again. Line 4 adds the newly failed nodes to the set of failed nodes. Line 5 enforces $f < n$. Since this action is only called once per round we do not need an extra relation \textit{failing\_now} presented in \textit{vii.1}.

@@

\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
invariant system.decision_round <->
            (forall N . (~node.failed(N) ->
                          (exists V . node.decision(N,V))))
\end{lstlisting}
\end{mdframed}
This invariant assumes that the system will eventually reach the decision round. However, truly verifying \textit{termination} would require checking that the system must reach the decision round, which means verifying  \textit{liveness}. IVy theoretically allows \textit{liveness} checks by constructing a reduction of \textit{liveness} to \textit{safety} \cite{live} but this is out of the scope of this work.

\subsection{Invariants for Inductiveness and IVy's CTI-Viewer}
If one were to run $ivy\_check$ on the program without specifying any additional invariants, IVy finds that neither the \textit{agreement} invariant, nor the \textit{validity} invariant is inductive. However, this does not immediately imply that \textit{Floodset}, or our specification of it, is incorrect. It means that the specified system can transition from a state in compliance with invariants to a state that is not. Thus, to prove \textit{Floodset} we need to find additional invariants that exclude unreachable states so that IVy can proof induction of the invariants that were found so far.

To improve the readers' intuition for this process it helps to (wrongly) think of IVy as a system searching the state space for counterexamples to induction as follows:
\begin{enumerate}
  \item Choose an initial system state, i.e. assignments to types, relations and functions, that is compliant with all invariants.
  \item Execute an exported action which mutates the initial system state.
  \item Check if the system state after the execution violates any of the invariants, if yes, display counterexample to induction (CTI).
  \item Goto 1.
\end{enumerate}

While this is not how IVy works internally, it illustrates an important point made in \textit{section I}: IVy does not consider the reachability of a system state. As long as a state is in compliance with the conjunction of all specified invariants it is assumed to be reachable. When IVy finds a counterexample to induction (CTI), it means that the system can transition from a state in compliance with invariants to a state that is not. It does not however imply, that there exists an execution path of exported actions so that the system, starting in its initial state defined in the \textit{init} action, can transition to a state that violates an invariant. We therefore have to strengthen our invariants, i.e. limit the state space that IVy can use to find CTIs, by adding invariants that exclude unreachable system states.

As this is not a straightforward process, IVy uses a GUI to illustrate CTIs to the user. If, like in our case, multiple invariants fail the check for inductiveness, IVy draws a counterexample to the invariant that is defined earliest in the specification. For us, IVy shows a counterexample to \textit{agreement}.

@@

\noindent In the left hand section of the program we can see, that the system transitioned from state 0
to state 1 by executing \textit{round}. In the center IVy displays state 0, meaning the state
before \textit{round} was executed. This state is in compliance with the set of invariants.
The brown edges specify the relation $w$, the purple edges show $val$.
Clicking on state 1 would show a state that violates the \textit{agreement} invariant.

The user must now understand why state 0 is unreachable.
Observing the graph of state 0, one can see that no node has their initial value, which is modeled by the relation $val$,
in its $w$. The user must now exclude this state by coming up with an invariant.

@@

After adding this invariant, IVy can no longer construct a CTI with a state 0 that violates the added invariant. However, even after adding this invariant, IVy still fails to prove induction of \textit{agreement} and \textit{validity} as it still assumes a variety of unreachable states to be reachable.

Thus, to prove \textit{Floodset}, the user must iterate the presented method for finding the remaining invariants: Execute $ivy\_check$ $\rightarrow$ examine CTI $\rightarrow$ design invariant $\rightarrow$ execute $ivy\_check$.

The following snippets presents the necessary invariants for induction annotated with comments.
\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
# initial value is in W
invariant node.w(N, node.val(N))

# the global default is the one the nodes know
invariant node.default(N) = node.def

@@

The most complex of these invariants is on line 12, which is equivalent to Lemma 1 presented in \textit{section III}. The necessity to include such an invariant in the final proof shows that automatic proving in IVy still requires fundamental understanding of the algorithm. While the CTI-GUI helps users understand individual counterexamples, it does not come up with invariants on its own. It is therefore not a substitute for knowledge about the fundamental ideas behind an algorithm or protocol.

After these four invariants are added to the specification, IVy can successfully prove induction of the system. As the specified invariants entail the correctness conditions, the correctness of \textit{Floodset} has been proven.

@@

Since any check on \textit{Floodset}'s correctness assumes that the \textit{framework} has been modeled correctly, it makes sense to include correctness conditions specifically aimed at the \textit{framework} in the form of invariants. These invariants are not necessary for induction. Removing them will not cause any of IVy's other checks to fail. Their sole purpose is to give the user confidence that the \textit{framework} is specified correctly.

@@

\section{Evaluation}
How viable is IVy as a tool for algorithm designers proving correctness
of their work? The goal, when turning to a tool like IVy, is certainty of
correctness. However, due to a variety of reasons, absolute certainty is never achieved.
This section evaluates IVy's usefulness pointing out strengths and limitations that IVy users should keep in mind
when working with the system.

\subsection{Equivalence of specification and algorithm}
One can not trivially determine, that the specification in the IVy language is logically equivalent to the actual algorithm the user wishes to verify. Whether the algorithm to be verified exists in the user's mind, in the form of pseudo-code or fully implemented in a programming language, no guarantee can be given in regards to the correctness of the translation of the algorithm into the IVy language. Some examples illustrating this difficulty will be given in the following paragraphs.

\subsubsection{Structural difference of IVy and procedural languages}
The IVy language is structurally unlike classical procedural languages. This makes it inherently challenging to achieve a one-to-one translation from a classical distributed algorithm, that was specified in a procedural way, like \textit{Floodset}. An example of this structural difference is the relatively complicated quantifier-logic in assignments that update the global network state. Take as an example the first line of \textit{step2}:

@@

This comparison clearly shows the complex structural differences. While the pseudo-code is concerned only with state alteration of a single node, the IVy language code simultaneously alters the state of all nodes. It uses an existential quantifier nested within a term. Its logic touches on multiple layers of abstraction as it mixes system and node concerns. It considers node failure, message reception and node state alteration.

As these two lines are so different, equivalence may not be obvious to a user. Proving equivalence of pseudo-code and IVy language specification would be a non-trivial proof exercise that certainly trumps manually proving \textit{Floodset}'s correctness in difficulty.

@@

\subsection{Accidental invalidation of legal state configurations with \textit{assume}}
$assume$ statements may silently invalidate state configurations. They therefore pose a threat as they
may allow false-positive IVy checks.
To give a practical example, the author of this paper originally modeled the line

@@

\noindent For example, one may insert
\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
assert ~(exists N . node.failed(N) &
        (exists J . ~incoming_message(J,N) &
          exists K . incoming_message(K,N) & J~=N & K~=J & N~=K))
\end{lstlisting}
\end{mdframed}
after the \textit{simulate\_failure} action. The negated part, $\gamma$, specifies that there exists a node $N$, that failed and failed to deliver a message to $J$ while successfully delivering a message to $K$, with $N,J,K$ distinct. By negating $\gamma$, we force IVy to construct a counterexample such that $\gamma$ is satisfied. If IVy does not find a counterexample to this assertion, it means that $\gamma$ cannot occur. In this particular case, that would mean the specification does not allow partial message delivery. The specification would therefore be erroneous. One has to finally remove the failing assertion to allow IVy to prove induction again.

@@

\subsection{Necessity of deep understanding}
It has become apparent, that successful verification with IVy is not possible without deep knowledge about how the algorithm functions. IVy is not a ``translate and go'' solution.
Evidence of this include the complex invariant that requires knowledge about \textit{lemma 1} or the alternative encoding for $f<n$ and the decision condition $if\ rounds = f+1$.
